\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-03-29 paper.qmd
done with first part of section 3

<!-- - Connection between NLDR and MDS-->
Historically, $k-D$ representations of $p-D$ data have been computed using multidimensional Scaling (MDS), which includes principal components analysis (PCA) as a special case.  The $k-D$ representation can be considered to be a layout of points in $k-D$ produced by an embedding procedure that maps the data from $p-D$. In MDS, the $k-D$ layout is constructed by minimizing a stress function that differences distances between points in $p-D$ with potential distances between points in $k-D$. Various formulations of the stress function result in non-metric scaling [@article62] and isomap [@article63]. Challenges in working with high-dimensional data, including visualization, are outlined in @Johnstone2009.

@@

@@

>>>>

<!-- - Connection between NLDR and MDS-->
Historically, $k-D$ representations of $p-D$ data have been computed using multidimensional scaling (MDS), which includes principal components analysis (PCA) as a special case.  The $k-D$ representation can be considered to be a layout of points in $k-D$ produced by an embedding procedure that maps the data from $p-D$. In MDS, the $k-D$ layout is constructed by minimizing a stress function that differences distances between points in $p-D$ with potential distances between points in $k-D$. Various formulations of the stress function result in non-metric scaling [@article62] and isomap [@article63]. Challenges in working with high-dimensional data, including visualization, are outlined in @Johnstone2009.

@@

At first glance, thinking of of NLDR as a model fitted to the data might seem strange. It is a model in the sense that it is a "a simplified representation or abstraction of a system, process, or phenomenon in the real world" where the $p-D$ observations are the realization. From a statistical perspective we can consider the distances between points in the $k-D$ layout to be variance that the model explains, and the (relative) difference with their distances in $p-D$ is the error, or unexplained variance. Abstractly, we can also imagine that the positioning of points in 2D represents fitted values, that will have some prescribed position in $p-D$ that can be compared with the observed value.

@@

<< XXX Jayani, please insert a table of notation >>

Once we have notation, we need to have some math that gives precision to the paragraph above. (XXX Michael??)

### 2024-03-29 paper.tex
done with first part of section 3

Historically, \(k-D\) representations of \(p-D\) data have been computed
using multidimensional Scaling (MDS), which includes principal
components analysis (PCA) as a special case. The \(k-D\) representation
can be considered to be a layout of points in \(k-D\) produced by an
embedding procedure that maps the data from \(p-D\). In MDS, the \(k-D\)
layout is constructed by minimizing a stress function that differences
distances between points in \(p-D\) with potential distances between
points in \(k-D\). Various formulations of the stress function result in
non-metric scaling \citep{article62} and isomap \citep{article63}.
Challenges in working with high-dimensional data, including
visualization, are outlined in \citet{Johnstone2009}.

@@

@@

>>>>

Historically, \(k-D\) representations of \(p-D\) data have been computed
using multidimensional scaling (MDS), which includes principal
components analysis (PCA) as a special case. The \(k-D\) representation
can be considered to be a layout of points in \(k-D\) produced by an
embedding procedure that maps the data from \(p-D\). In MDS, the \(k-D\)
layout is constructed by minimizing a stress function that differences
distances between points in \(p-D\) with potential distances between
points in \(k-D\). Various formulations of the stress function result in
non-metric scaling \citep{article62} and isomap \citep{article63}.
Challenges in working with high-dimensional data, including
visualization, are outlined in \citet{Johnstone2009}.

@@

At first glance, thinking of of NLDR as a model fitted to the data might
seem strange. It is a model in the sense that it is a ``a simplified
representation or abstraction of a system, process, or phenomenon in the
real world'' where the \(p-D\) observations are the realization. From a
statistical perspective we can consider the distances between points in
the \(k-D\) layout to be variance that the model explains, and the
(relative) difference with their distances in \(p-D\) is the error, or
unexplained variance. Abstractly, we can also imagine that the
positioning of points in 2D represents fitted values, that will have
some prescribed position in \(p-D\) that can be compared with the
observed value.

@@

\textless\textless{} XXX Jayani, please insert a table of notation
\textgreater\textgreater{}

Once we have notation, we need to have some math that gives precision to
the paragraph above. (XXX Michael??)
